Insight Into Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-08-25 draft.tex
Results figure now fits its canvas
filling blank spots

\section{\uppercase{Experiments}}
\label{sec:Experiments}
\noindent Spine shape is confounded by several somatometric variables.
%%
Larger people have also a longer spine and its shape is more straight compared to smaller subjects.
%%
Since men are on average taller than women, gender is another confounder.
%%
Large body weight increases the spine load, resulting in a bend shape.
%%
All these variables need to be taken into account, when correlating spine curvature and torsion with non-image parameter.
%%
Since the gender confounder mainly encodes body-height, we decided to divide subjects into body-size groups.
%%
To avoid small outlier groups, epidemiologists recommend using quartiles to discretize metric variables \cite{KlemmVIS14}.
\subsection{Preliminary Results}
\begin{figure*}[htb]
  %\vspace{-0.2cm}
  \centering
  \includegraphics[width=1.0\textwidth]{figures/image-parameter-range}
  \caption{image-parameter-range}
  \label{fig:image-parameter-range}
\end{figure*}
%%
\noindent As first experiment we correlated the shape parameter with the binary back pain indicator using a pairs plot \cite{ggally}.
%%
Pairs plots visualize continuous and categorical variables pairwise in a matrix view.
%%
Scatterplots display pairs of continuous variables, bar charts for continuous variable as a function of a categorical variable and box plots pairs of categorical variables.
%%
Since our image-derived variables are metric, their pairwise combinations are visualized using scatterplots.
%%
The combination of the image variables with back pain is visualized as histogram at the left side of the matrix and as box plot on the right side.
%%
Figure~\ref{fig:image-parameter-range} (right) shows the range of each variable as box plot.
%%
The projections to the transversal planes attract attention as they have many outliers.
%%
We conclude that curvature is not as reliable on the transversal plane as it is on the other planes.
%%--------------------------------
\subsubsection{Correlation Matrix}
We calculated an association matrix to to assess correlations between the image parameters.
%%
The pearson correlation coefficients between the numeric variables are depicted right of the main diagonal of Figure~\ref{fig:image-parameter-range}.
%%
Curvature, curvature angle and torsion correlate strongly with their planar projections.
%%
Also the mean curvature and the curvature angle correlate by a factor of -0.89.
%%
%Torsion and curvature/curvature angle do not correlate.
Torsion does not correlate with any other image-derived variable.
%%--------------------------------------------------------
\subsubsection{Correlation of Image Parameters With Back Pain}
Figure~\ref{fig:image-parameter-range} shows the distribution for all subjects.
%%
We created a pairs plot for subject groups derived by dividing them into quantile groups of body-size.
%%
No statistical significant correlation could be observed through all subject groups.
%%
The box plots show no difference between subjects with and without back pain.

\subsubsection{Assessing the Information Gain Using the PCA}
%%
To determine the information gain per image-derived variables, we calculated a PCA and compared the loadings per dimension.
%%
The first three principal components explain 75\% of variance in the image-derived variables.
%%
The first principal component explains 47\% of the variance and weights primarily \emph{mean curvature}, \emph{curvature sagittal}, \emph{curvature angle} and \emph{curvature angle sagittal}.
%%
The second component, adding 16\% of variance, weigths  \emph{curvature coronal} and \emph{curvature angle coronal}.
%%
The third component explains 12\% of variance and weights \emph{torsion} and \emph{curvature transverse}.
%%
This supports our prior conclusion about the low information gain of the transverse planes.
%%
\emph{Torsion} also adds little variance to the information space.
%% ------------------------------------------
\subsubsection{Heterogenous Correlations}
%%
We then expanded our focus on correlations of image-derived parameters with all other non-image variables.
%%
We applied a heterogenous correlation technique to derive correlations between all variables is the data set.
%%
The method uses the following correlation metrics for the different type combination:
%%
\begin{itemize}
%%
\item \emph{pearson product-moment} for two continuous variables, 
\item \emph{polyserial} correlation for one continuous and one categorical variable, and
\item \emph{polychoric} correlations for two categorical variables.
\end{itemize}
%%
All correlation values are scaled between 0 - no correlation and 1 - identical.
%%
Some variables are too sparse for calculating correlations, for example \com{TODO: some variables from calculation here; TODO: Image?, eventuell interactiven Tile-Plot}.
%%
We display the resulting \emph{contingency matrix} using a heat map, mapping correlation values to color brightness with white for 0 and dark blue for 1 \cite{Klemm2014VIS}.
%%
We calculated the contingency matrix for all size groups and looked for correlations between image- and non-image derived variables.
%%
The resulting contingency matrix shows no strong correlation with any of the parameter.
%%
Only weak correlations could be found for \emph{Mean Curvature} with \emph{gender} (0.42), \emph{body size} (0.39) and \emph{number of born children} (0.29).
%%
One surprising result was the small correlation with \emph{torsion}, which correlated with almost no variables (p values between 0 and 0.05) and \emph{parkinson} (0.24).
%%
This observation brought us to the conclusion to incorporate more sophisticated data mining techniques to assess the influence of the image-derived parameters.

@@

\subsection{Interactive Display of Decision Trees}
We have to create a decision tree for every non-image variable to see which one can be explained by image-derived variables.
%%
Since we have 134 non-image variables, the calculation yields a corresponding number of trees.
%%
Dividing the subjects into four groups using body-size, the number increases to 402 trees.
%%
To make the results cognitive feasible, we have to abstract the results of the classification.
%%------------
\subsubsection{Visualization of Classification Results}
%%
We follow the visual information-seeking mantra and want to provide an overview first, then details-on-demand \ref{shneiderman1996}.
%%
The optimal classification uses a few rules to precisely characterize the target variable.
%%
Therefore, we are interested in \emph{small trees} with a \emph{low classification error rate}.
%%
The two measures form the axis for a \emph{scatter plot} of the classification results.

@@

Many follow-up variables are sparse, such as medication of diabetes or reason of early retirement.
%%
The classification algorithm may produce higher accuracy for variables with less subjects due to the small sample size.
%%
This makes these results less reliable.
%%
We map the number of subjects associated with a variable to point diameter in the scatter plot.
%%
This allows to instantly assess the reliability of the result.
%%

@@

\com{Needs to be moved}
\add{We removed variables with less than 100 subjects, since they have to few subjects to be statistically significant.}
\com{ToDo: Manche Variablen rausgeschmissen, weil sie zu wenig Probanden enthielten, oder Auspraegungen nur duenn besetzt waren!}

%\com{ToDo: SQRT Skalierung der Treesize-Dimension, da hauptsaechlich geringe Groessen von interesse sind, ausreisser aber diesen Bereich schwer durchschaubar machen}
% %%
%
% %%
% Abstracting the classification results into two continuos variables allows us to visualize the result of 134 classifications in one scatter plot.
% %%
% By mapping the group affiliation (in our case body-size) to color or shape, we can even plot all 402 results in a single visualization (Fig.~\ref{fig:decision-tree} a).
%%
\subsubsection{Dummy Variables}
Dummy variables abstract a categorical variable with multiple manifestations into several binary variables.
%%
Each binary variables encodes the presence of a manifestation.
%%
For example, a pain indicator variable ranging from \emph{1 - no pain} to \emph{4 - large pain} is subdivided into four binary variables (\emph{No pain - Yes/No} to \emph{Large Pain - Yes/no}).
%%
One subject can only have one of these variables set to true.
%%
This is useful for our classification, because it allows to determine which manifestation of a variable can be described best using the image data parameter.
%%\com{ToDo: Dummy-Variables}
%%------------
\subsubsection{Interaction With the Visualization}
%%
The described visualization provides a good overview over the classification results.
%%
We however still want to be able to look display \emph{details-on-demand} \ref{shneiderman1996} and examine a decision tree in detail.
%%
This is realized by clicking on a entry on the visualization, which then displays the corresponding decision tree in detail (Fig.~\ref{fig:decision-tree} b).
%%
This allows for sequentially analyzing the classifications.
%%------------
\subsubsection{Implementation}
All analysis are carried out using \texttt{R}, a popular programming language for statistical calculations and visualizations.
%%
The interactive visualizations are realized using the \texttt{ggvis}\footnote{GGVIS Footnote} package.
%%
As opposed to the standard \texttt{R} plots, \texttt{ggvis} allows to adjust visualization parameters using UI controls, such as sliders.
%%
We implemented our application as web based prototype using \texttt{R Shiny}\footnote{R Shiny footnote}.
%%
The package combines the power of \texttt{R} with advantages of web-based applications.
%%
It allows us to quickly exchange results with our collaborating epidemiological partners.
%%
They can try out the technique without installing anything.
%%
Exchanging software becomes as easy as exchanging a hyperlink.
%%
\com{TODO: Auf Eignung von RMarkdown als Report eingehen - ergebnisse koennen schnell publiziert und mit neuen Daten durchgetestet werden.}
%%---------
\subsection{Results}
\begin{figure*}[p!]
  %\vspace{-0.2cm}
  \centering
  \includegraphics[width=1.0\textwidth]{figures/results}
  \caption{myResults}
  \label{fig:results}
\end{figure*}
We ran the analysis using different subject groups:
\begin{enumerate}
	%%
	\item All subjects,
	%%
	\item subdivision into \emph{male} and \emph{female},
	%%
	\item subdivision into \emph{age quantiles}, yielding the groups (21, 43.8] (43.8, 54] (54, 64] (64, 90],
	%%
	\item subdivision into \emph{size quantiles}, yielding the groups (139, 164], (164, 171], (171, 177], (177, 202].
	%%
\end{enumerate}
We plotted each group twice.
%%
The first plot shows all original variables.
%%
The second plot shows all categorical variables transformed into dichotomous dummy variables.

@@

%\com{Potentially skip these due to the confounding of body size}
\subsubsection{Age Groups}
%%
Gender could be described for each age group using \emph{mean curvature} and \emph{curvature angle}.
%%
The accuracy varies between $0,35$ ($43,8-54$ years of age, 3 decision rules) to $0,27$ ($21-43,8$ years of age, 6 decision rules).
%%
Many variables, such as body size can be described with an error rate of $0,35$ to $0,45$ but only using large decision trees with over 20 rules.
%Body-size can be determined with an error rate of $0,36$ to $0,39$ for subject
% age groups, but also imposes over fitting to the data yielding a decision tree of over 30 rules.
%%
Notable is also the increasing accuracy with increasing decision tree size.
%%
For subjects between $43,8-54$ years of age, the classifier discriminated subjects with thyroid nodules with an error rate of $0,32$.

@@

\com{Manchmal kann es passieren, dass die Quantile fuer eine Variable gleich sind - z.B. bei Dauer Jahre Schwerarbeit, weil viele Probanden hier eben eine 0 haben. Deswegen kann das Ding nicht in vier Qunatile unterteilt werden, die ersten 3 sind 0. Deswegen wurden nur die einzigartigen Unterscheidungen genommen (statt \emph{quantile(myVar)} wurde \emph{unique(quantile(myVar))} verwendet.)}

\subsection{Themen hier unterbringen}

\com{We can discriminate back pain using non-image data, but not with image-derived parameter. Welche parameter koennen wir gut unterscheiden}

\com{Allgmeine Analyse der Daten mit einbringen. Hierfuer statistisch Auswertung machen, welche Parameter am meisten verwendet wurden bei der Kategorisierung. Nur Parameter verwenden, die genauer sind als 0.5\%, der Entscheidungsbaum kleiner als 15 ist und das Feature mindestens 100 Probanden enthält}

\com{Weiterhin interessant die Beschreibung rueckenschmerzassoziitierter Parameter, bei einigen zeigt sich, dass Torsion sehr dominant vertreten ist.}

\com{Algorithmus schreiben, der eng zusammenliegende Parameter fuer alle Probanden sucht.}

>>>>

\section{\uppercase{Experiments}}
\label{sec:Experiments}
\noindent Spine shape is confounded by several somatometric variables.
%%
Larger people have also a longer spine and its shape is more straight compared to smaller subjects.
%%
Since men are on average taller than women, gender is another confounder.
%%
Large body weight increases the spine load, resulting in a bend shape.
%%
All these variables need to be taken into account, when correlating spine curvature and torsion with non-image parameter.
%%
Since the gender confounder mainly encodes body-height, we decided to divide subjects into body-size groups.
%%
To avoid small outlier groups, epidemiologists recommend using quartiles to discretize metric variables \cite{KlemmVIS14}.
\subsection{Preliminary Results}
\begin{figure*}[htb]
  %\vspace{-0.2cm}
  \centering
  \includegraphics[width=1.0\textwidth]{figures/image-parameter-range}
  \caption{%%
A generalized pairs plot of all image-derived variables \cite{GPLOMS} colored by presence or absence of back pain.
%%
Pairwise combinations of image-derived variables are visualized using scatter plots left of the matrix diagonal and their correlation with back pain is denoted right of the matrix diagonal.
%%
The box plots (left) and histograms (bottom) display the distribution of each image-variable encoded with back pain.
%%
No correlation with back pain are present can be seen in this plot.
}
  \label{fig:image-parameter-range}
\end{figure*}
%%
\noindent As first experiment we correlated the shape parameter with the binary back pain indicator using a generalized pairs plot \cite{GPLOMS}.
%%
Pairs plots visualize continuous and categorical variables pairwise in a matrix view.
%%
Scatterplots display pairs of continuous variables, bar charts for continuous variable as a function of a categorical variable and box plots pairs of categorical variables.
%%
Since our image-derived variables are metric, their pairwise combinations are visualized using scatterplots.
%%
The combination of the image variables with back pain is visualized as histogram at the left side of the matrix and as box plot on the right side.
%%
Figure~\ref{fig:image-parameter-range} (right) shows the range of each variable as box plot.
%%
The projections to the transversal planes attract attention as they have many outliers.
%%
We conclude that curvature is not as reliable on the transversal plane as it is on the other planes.
%%--------------------------------
\subsubsection{Correlation Matrix}
We calculated an association matrix to to assess correlations between the image parameters.
%%
The pearson correlation coefficients between the numeric variables are depicted right of the matrix diagonal in Figure~\ref{fig:image-parameter-range}.
%%
Curvature, curvature angle and torsion correlate strongly with their planar projections.
%%
Also the mean curvature and the curvature angle correlate by a factor of -0.89.
%%
%Torsion and curvature/curvature angle do not correlate.
Torsion does not correlate with any other image-derived variable.
%%--------------------------------------------------------
\subsubsection{Correlation of Image Parameters With Back Pain}
Figure~\ref{fig:image-parameter-range} shows the distribution of all image-parameters \add{as generalized pairs plot}.
%%
\rem{We created a generalized pairs plot for subject groups derived by dividing them into quantile groups of body-size.}
%%
No statistical significant correlation could be observed through all subject groups.
%%
The box plots show no difference between subjects with and without back pain.

\subsubsection{Assessing the Information Gain Using the PCA}
%%
To determine the information gain per image-derived variables, we calculated a PCA and compared the loadings per dimension.
%%
The first three principal components explain 75\% of variance in the image-derived variables.
%%
The first principal component explains 47\% of the variance and weights primarily \emph{mean curvature}, \emph{curvature sagittal}, \emph{curvature angle} and \emph{curvature angle sagittal}.
%%
The second component, adding 16\% of variance, weigths  \emph{curvature coronal} and \emph{curvature angle coronal}.
%%
The third component explains 12\% of variance and weights \emph{torsion} and \emph{curvature transverse}.
%%
This supports our prior conclusion about the low information gain of the transverse planes.
%%
\emph{Torsion} also adds little variance to the information space.
%% ------------------------------------------
\subsubsection{Heterogenous Correlations}
%%
We then expanded our focus on correlations of image-derived parameters with all other non-image variables.
%%
We applied a heterogenous correlation technique to derive correlations between all variables is the data set.
%%
The method uses the following correlation metrics for the different type combination:
%%
\begin{itemize}
%%
\item \emph{pearson product-moment} for two continuous variables, 
\item \emph{polyserial} correlation for one continuous and one categorical variable, and
\item \emph{polychoric} correlations for two categorical variables.
\end{itemize}
\com{TODO: Abbildung einer Korrelationsmatrix? Waere sehr gross!}
%%
All correlation values are scaled between 0 - no correlation and 1 - identical.
%%
Some variables are too sparse for calculating correlations, for example \emph{treatment of diabetes}, or \emph{medication against high blood pressure}.
%%
We display the resulting \emph{contingency matrix} using a heat map, mapping correlation values to color brightness with white for 0 and dark blue for 1 \cite{Klemm2014VIS}.
%%
We calculated the contingency matrix for all size groups and looked for correlations between image- and non-image derived variables.
%%
The resulting contingency matrices are available as supplementary material and show no strong correlation with any of the parameter.
%%
Only weak correlations could be found for \emph{Mean Curvature} with \emph{gender} (0.42), \emph{body size} (0.39) and \emph{number of born children} (0.29).
%%
One surprising result was the small correlation with \emph{torsion}, which correlated with almost no variables (p values between 0 and 0.05) and \emph{parkinson} (0.24).
%%
This observation brought us to the conclusion to incorporate more sophisticated data mining techniques to assess the influence of the image-derived parameters.

@@

\subsection{Interactive Display of Decision Trees}
We have to create a decision tree for every non-image variable to see which one can be explained by image-derived variables.
%%
Since we have 134 non-image variables, the calculation yields a corresponding number of trees.
%%
Dividing the subjects into four groups using body-size, the number increases to 402 trees.
%%
To make the results cognitive feasible, we have to abstract the results of the classification.
%%------------
\subsubsection{Visualization of Classification Results}
\label{subsec:VisualizationOfClassificationResults}
%%
We follow the visual information-seeking mantra and want to provide an overview first, then details-on-demand \ref{shneiderman1996}.
%%
The optimal classification uses a few rules to precisely characterize the target variable.
%%
Therefore, we are interested in \emph{small trees} with a \emph{low classification error rate}.
%%
The two measures form the axis for a \emph{scatter plot} of the classification results.

@@

Many follow-up variables are sparse, such as medication of diabetes or reason of early retirement.
%%
The classification algorithm may produce higher accuracy for variables with less subjects due to the small sample size.
%%
This makes these results less reliable.
%%
We removed variables with less than 100 subjects, since they have to few subjects to be statistically significant.
%%
Furthermore, we map the number of subjects associated with a variable to point diameter in the scatter plot.
%%
This allows to instantly assess the reliability of the result.
%%

@@

%\com{ToDo: SQRT Skalierung der Treesize-Dimension, da hauptsaechlich geringe Groessen von interesse sind, ausreisser aber diesen Bereich schwer durchschaubar machen}
% %%
%
% %%
% Abstracting the classification results into two continuos variables allows us to visualize the result of 134 classifications in one scatter plot.
% %%
% By mapping the group affiliation (in our case body-size) to color or shape, we can even plot all 402 results in a single visualization (Fig.~\ref{fig:decision-tree} a).
%%
\subsubsection{Dummy Variables}
Dummy variables abstract a categorical variable with multiple manifestations into several binary variables.
%%
Each binary variables encodes the presence of a manifestation.
%%
For example, a pain indicator variable ranging from \emph{1 - no pain} to \emph{4 - large pain} is subdivided into four binary variables (\emph{No pain - Yes/No} to \emph{Large Pain - Yes/no}).
%%
One subject can only have one of these variables set to true.
%%
This is useful for our classification, because it allows to determine which manifestation of a variable can be described best using the image data parameter.
%%\com{ToDo: Dummy-Variables}
%%------------
\subsubsection{Interaction With the Visualization}
%%
The described visualization provides a good overview over the classification results.
%%
We however still want to be able to look display \emph{details-on-demand} \ref{shneiderman1996} and examine a decision tree in detail.
%%
This is realized by clicking on a entry on the visualization, which then displays the corresponding decision tree in detail (Fig.~\ref{fig:decision-tree} b).
%%
This allows for sequentially analyzing the classifications.
%%------------
\subsubsection{Implementation}
All analysis are carried out using \texttt{R}, a popular programming language for statistical calculations and visualizations.
%%
The interactive visualizations are realized using the \texttt{ggvis}\footnote{GGVIS Footnote} package.
%%
As opposed to the standard \texttt{R} plots, \texttt{ggvis} allows to adjust visualization parameters using UI controls, such as sliders.
%%
In order to make the train of thought comprehensible, we used \texttt{RMarkdown}, which allows to make reports by combining \texttt{R} with the \texttt{Markdown} syntax.
%%
We used \texttt{R Shiny}\footnote{R Shiny footnote} to make the report available as dynamic web application.
%%
It allows to combine both the ability to create comprehensible reports as well as UI-controls, to adjust visible data and parameters in the plot to allow the viewer to analyze the data.
%%
Furthermore, calculations based on a prior data selection can be redone within the report.
%We implemented our application as web based prototype using \texttt{R Shiny}\footnote{R Shiny footnote} in combination with \texttt{RMarkdown}.
%%
%The package combines the power of \texttt{R} with advantages of web-based applications.
%%
The web-based approach allows us to quickly exchange results with our collaborating epidemiological partners.
%%
They can try out the technique without installing any software.
%%
Exchanging the prototype becomes as easy as exchanging a hyperlink.
%%---------
\subsection{Results}
\begin{figure*}[p!]
  %\vspace{-0.2cm}
  \centering
  \includegraphics[width=0.8\textwidth]{figures/results}
  \caption{
%%
Scatterplot visualization of all C5.0 classification results.
%%
The x-axis shows the number of decisions of the underlying model, the y-axis the classification error rate (see Section~\ref{subsec:VisualizationOfClassificationResults}).
%%
The left scatterplot shows the results for all variables, either metrics expressed via their quantiles, or categorical.
%%
The right scatterplot displays the dummy variables derived from the original variables.
%%
Group affiliation of a data point is color coded:
%%
no group (a), subdivision into male and female subjects (b), quartiles of age (c) and quartiles of body size (d).
%%
%(a) displays all subjects, (b) is divided into male and female subjects, (c) is categorized into quantiles of age and (d) shows the classification of subjects divided by quantiles of body size.
%%
The number of subjects represented in a variable is denoted using the dot diameter.
%%
We only display variables with an error rate below $0,5$, results above this threshold are inaccurate.
%%
A interactive versions of the plot (see supplemental material) makes all data points clickable, showing the corresponding decision tree in a tool tip.
%%
}
  \label{fig:results}
\end{figure*}
We ran the analysis using different subject groups:
\begin{enumerate}
	%%
	\item All subjects,
	%%
	\item subdivision into \emph{male} and \emph{female},
	%%
	\item subdivision into \emph{age quantiles}, yielding the groups (21, 43.8] (43.8, 54] (54, 64] (64, 90],
	%%
	\item subdivision into \emph{size quantiles}, yielding the groups (139, 164], (164, 171], (171, 177], (177, 202].
	%%
\end{enumerate}
We plotted each group twice.
%%
The first plot shows all original variables.
%%
The second plot shows all categorical variables transformed into dichotomous dummy variables.

@@

%\com{Potentially skip these due to the confounding of body size}
\subsubsection{Age Groups}
%%
Gender could be described for each age group using \emph{mean curvature} and \emph{curvature angle}.
%%
The accuracy varies between $0,35$ ($43,8-54$ years of age, 3 decision rules) to $0,27$ ($21-43,8$ years of age, 6 decision rules).
%%
The starting age of smoking could be characterized well with an accuracy between $0,25$ to $0,3$ for all age groups except of $21,1-43,8$ years of age.
%%
The result however is probably over fitted to the data due to tree sizes between $16$ and $23$.
%%
\com{Feedback der Greifswalder notwendig, warum das so sein koennte.}
%%

Many variables, such as body size can be described with an error rate of $0,35$ to $0,45$ but only using large decision trees with over 20 rules.
%Body-size can be determined with an error rate of $0,36$ to $0,39$ for subject
% age groups, but also imposes over fitting to the data yielding a decision tree of over 30 rules.
%%
Notable is also the increasing accuracy with increasing decision tree size.
%%
For subjects between $43,8-54$ years of age, the classifier discriminated subjects with thyroid nodules with an error rate of $0,32$.

@@

%TODO: Muss das wirklich noch rein?
% \com{Manchmal kann es passieren, dass die Quantile fuer eine Variable gleich sind - z.B. bei Dauer Jahre Schwerarbeit, weil viele Probanden hier eben eine 0 haben. Deswegen kann das Ding nicht in vier Qunatile unterteilt werden, die ersten 3 sind 0. Deswegen wurden nur die einzigartigen Unterscheidungen genommen (statt \emph{quantile(myVar)} wurde \emph{unique(quantile(myVar))} verwendet.)}

\com{Auswertung der Anteile der Parameter an den Ergebnissen (Torsion, Curvature, etc. - welches ist der dominanteste Parameter?)}
